Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-07-30 submission/manuscript.Rmd
update methods section

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (n=`r length(metaf$initial)`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

@@

**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].

@@

**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification models [@randomforest_citation_2002]. This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. The classification models were built to group normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Walis rank sum test with a Dunn's post-hoc test was used to assess differences between the groups used. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumor using Random Forest models. Next, the imputed gene counts of important mediators of SCFA synthesis was tested. Additionally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. Finally, models to classify high or low SCFA concentration based on the median of each SCFA and 16S rRNA gene sequencing data was created using the Random Forest algorithm.

>>>>

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =`r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())` and carcinoma n = `r length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

@@

**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identifed and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].

@@

**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification and regression models via the caret package [@randomforest_citation_2002; @caret_citation]. This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. The classification models were built to group normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations. The regression models were built to classify the SCFA concentrations of acetate, butyrate, and propionate regardless of disease status.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Walis rank sum test with a Dunn's post-hoc test was used to assess differences between the groups used. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, the imputed gene counts of important mediators of SCFA synthesis was tested. Additionally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without tumors and patients with adenomas or carcinomas using Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models. Finally, models to classify high or low SCFA concentration based on the median of each SCFA or the actual concentration using 16S rRNA gene sequencing data was created using the Random Forest algorithm. For all Random Forest models, the assessment of the most important variables was based on the top 10 feautres (OTUs or SCFAs) using the mean decrease in accuracy.

### 2018-07-30 submission/manuscript.tex
update methods section

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (29, 30). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (29).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (29, 30). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Declaration.

@@

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (29, 31, 32). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (33).

@@

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=490) containing a total of 85
individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (34). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (35). Open Reading Frames (ORFs) were
identified using Prodigal (36), counts generated using Diamond (37),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (38), and OPF alignment used the KEGG database (39).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (39,
40) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for classification models (21). This was
repeated on 100 different 80/20 splits of the data to generate a
reasonable range for the AUC of the model. The reported AUCs, unless
otherwise specified, are for the test sets. The classification models
were built to group normal versus adenoma, normal versus carcinoma, and
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (41). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (42). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA synthesis was tested.
Additionally, the counts generated for OPFs that matched important genes
involved with SCFA creation were analyzed. Finally, models to classify
high or low SCFA concentration based on the median of each SCFA and 16S
rRNA gene sequencing data was created using the Random Forest algorithm.

@@

\hypertarget{ref-opticlust_westcott2017}{}
33. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\hypertarget{ref-Hannigan2017}{}
34. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
\& the interactive dynamics of the colorectal cancer virome.
doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.

\hypertarget{ref-Li2015}{}
35. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
for large and complex metagenomics assembly via succinct de bruijn
graph. Bioinformatics \textbf{31}:1674--1676.
doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.

\hypertarget{ref-Hyatt2010}{}
36. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
Prodigal: Prokaryotic gene recognition and translation initiation site
identification. BMC Bioinformatics \textbf{11}:119.
doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.

\hypertarget{ref-Buchfink2014}{}
37. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
and sensitive protein alignment using DIAMOND. Nature Methods
\textbf{12}:59--60.
doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.

\hypertarget{ref-Steinegger2017}{}
38. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
sensitive protein sequence searching for the analysis of massive data
sets. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.

\hypertarget{ref-Kanehisa2015}{}
39. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.

\hypertarget{ref-scfa_baxter2014}{}
40. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-r_citation_2017}{}
41. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-benjamini_controlling_1995}{}
42. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society Series B
(Methodological) \textbf{57}:289--300.

>>>>

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (29, 30). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(adenoma n =41 and carcinoma n = 26). For patients undergoing treatment
for their tumor the length of time between their initial and follow up
sample ranged from 188 - 546 days. Our use of treatment has been
previously defined as encompassing removal of a tumor with or without
chemotherapy and radiation (29). Diagnosis of tumor was made by
colonoscopic examination and histopathological review of biopsies
obtained (29, 30). The University of Michigan Institutional Review Board
approved the study and informed consent was obtained from all
participants in accordance to the guidelines set out by the Helsinki
Declaration.

@@

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (29, 31, 32). In brief, sequences were quality
filtered and contigs created from the paired end reads. Any sequences
with ambiguous base calls were discarded. Contigs were then checked for
matches to the V4 region of the 16S rRNA gene using the SILVA database
(33). Chimeras were identifed and removed using UCHIME and OTUs
clustered at 97\% similarity (34). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (35).

@@

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=490) containing a total of 85
individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (36). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (37). Open Reading Frames (ORFs) were
identified using Prodigal (38), counts generated using Diamond (39),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (40), and OPF alignment used the KEGG database (41).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (41,
42) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for classification and regression models via the
caret package (21, 43). This was repeated on 100 different 80/20 splits
of the data to generate a reasonable range for the AUC of the model. The
reported AUCs, unless otherwise specified, are for the test sets. The
classification models were built to group normal versus adenoma, normal
versus carcinoma, and high versus low SCFA concentrations. The
regression models were built to classify the SCFA concentrations of
acetate, butyrate, and propionate regardless of disease status.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (44). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (45). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, the imputed gene counts of
important mediators of SCFA synthesis was tested. Additionally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. From here we analyzed the number of
significant positive and negative correlations between OTU relative
abundance and SCFA concentrations in individuals without tumors and
patients with adenomas or carcinomas using Spearman's rho. Next, we
assessed whether OTUs alone or OTUs and SCFAs were better able to
classify individuals with and without tumors using Random Forest models.
Finally, models to classify high or low SCFA concentration based on the
median of each SCFA or the actual concentration using 16S rRNA gene
sequencing data was created using the Random Forest algorithm. For all
Random Forest models, the assessment of the most important variables was
based on the top 10 feautres (OTUs or SCFAs) using the mean decrease in
accuracy.

@@

\hypertarget{ref-Quast2012}{}
33. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
\textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
J}, \textbf{Glöckner FO}. 2012. The SILVA ribosomal RNA gene database
project: Improved data processing and web-based tools. Nucleic Acids
Research \textbf{41}:D590--D596.
doi:\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.

\hypertarget{ref-Edgar2011}{}
34. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.

\hypertarget{ref-opticlust_westcott2017}{}
35. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\hypertarget{ref-Hannigan2017}{}
36. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
\& the interactive dynamics of the colorectal cancer virome.
doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.

\hypertarget{ref-Li2015}{}
37. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
for large and complex metagenomics assembly via succinct de bruijn
graph. Bioinformatics \textbf{31}:1674--1676.
doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.

\hypertarget{ref-Hyatt2010}{}
38. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
Prodigal: Prokaryotic gene recognition and translation initiation site
identification. BMC Bioinformatics \textbf{11}:119.
doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.

\hypertarget{ref-Buchfink2014}{}
39. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
and sensitive protein alignment using DIAMOND. Nature Methods
\textbf{12}:59--60.
doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.

\hypertarget{ref-Steinegger2017}{}
40. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
sensitive protein sequence searching for the analysis of massive data
sets. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.

\hypertarget{ref-Kanehisa2015}{}
41. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.

\hypertarget{ref-scfa_baxter2014}{}
42. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-caret_citation}{}
43. \textbf{Jed Wing MKC from}, \textbf{Weston S}, \textbf{Williams A},
\textbf{Keefer C}, \textbf{Engelhardt A}, \textbf{Cooper T},
\textbf{Mayer Z}, \textbf{Kenkel B}, \textbf{R Core Team},
\textbf{Benesty M}, \textbf{Lescarbeau R}, \textbf{Ziem A},
\textbf{Scrucca L}, \textbf{Tang Y}, \textbf{Candan C}, \textbf{Hunt.
T}. 2017. Caret: Classification and regression training.

\hypertarget{ref-r_citation_2017}{}
44. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-benjamini_controlling_1995}{}
45. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society Series B
(Methodological) \textbf{57}:289--300.
